---
title: "deterministic-models"
author: "Emily Guthrie"
date: "5/18/2020"
output: html_document
---

```{r}
library(deSolve)
library(plsgenomics)
library(tidyverse)
library(knitr)
```

From literature:
R0 is 2.2
duration for E is 3
duration for I is 7

# Basic SEIR model with no handwashing

Model
```{r}
OpenSEIR<-function(t, state, parameters) {
    with(as.list(c(state, parameters)),{
        N = S + E + I + R
        
        dS <- -beta*S*I/N + birth*N - death*S + omega*R
        dE <- beta*S*I/N - sigma*E - death*E
        dI <- sigma*E - death*I - gamma*I
        dR <- gamma*I - death*R - omega*R
        
        # return the rates of change as a list
        list(c(dS, dE, dI, dR))
    })
}

```

Parameters
```{r}
R0 <- 2.2
latent <- 3
infectious <- 7

birth <- 0
death <- 0
omega <- 0 #waning immunity
gamma <- 1/infectious #recovery rate
sigma <- 1/latent
beta <- R0*gamma #effective contact rate
```

Functions to analyze output
```{r}
compute_cumulative_infection_time<-function(output) {
  df1 <- data.frame(output)
  return(sum(df1$I))
}

show_SEIR_model_results<-function(df1) {
    df1 %>% as_tibble() %>% mutate(S = as.numeric(S), E = as.numeric(E), I = as.numeric(I), R = as.numeric(R)) %>%
    pivot_longer(cols = c("S", "E", "I", "R"), names_to = "compartment", values_to = "count") %>%
    mutate(compartment = factor(compartment, levels = c("S", "E", "I", "R"))) %>% 
    ggplot(aes(x = time, y = count, color = compartment)) +
        geom_line() +
        xlab('Time') +
        ylab('Count') 
}
```

Run model

```{r}
state <- c(S = 999999, #population of 1,000,000, 1 person starts of infected
           E = 0,
           I = 1, 
           R = 0
)

T_end <- 500 #run model for 500 time steps
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps, and computes output at each time step 

output_basic_seir <- ode(y = state, times = times, func = OpenSEIR, parms = c(beta, gamma, omega, sigma, birth, death))

output_basic_seir %>% show_SEIR_model_results()

compute_cumulative_infection_time(output_basic_seir)
```


# Basic SEIR with handwashing - affects infectious only

Model assuming handwashing only impacts infectious people

```{r}
handwash_I_SEIR<-function(t, state, parameters) {
    with(as.list(c(state, parameters)),{
        N <- S + E + I + R
        
        dW <- -theta*W + p*(N - W)
        
        dS <- -beta1*S*(1 - W/N)*(I/N) - beta2*S*(W/N)*(I/N) + birth*N - death*S + omega*R
        dE <- beta1*S*(1 - W/N)*(I/N) + beta2*S*(W/N)*(I/N) - sigma*E - death*E
        dI <- sigma*E - death*I - gamma*I
        dR <- gamma*I - death*R - omega*R
        
        # return the rates of change as a list
        list(c(dS, dE, dI, dR, dW))
    })
}
```

Parameters
```{r}
R0 <- 2.2
latent <- 3
infectious <- 7
lambda <- .2 #this is the percent by which handwashing reduces effective contact rate
p <- .2 #prob that someone washes hands
theta <- .2 #rate that handwashing wears off

birth <- 0
death <- 0
omega <- 0 #waning immunity
gamma <- 1/infectious #recovery rate
sigma <- 1/latent
beta1 <- R0*gamma #effective contact rate
beta2 <- (1 - lambda)*beta1
```

Run model

```{r}
state <- c(S = 999999, #population of 1,000,000, 1 person starts of infected
           E = 0,
           I = 1, 
           R = 0,
           W = 0
)

T_end <- 500 #run model for 500 time steps
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps, and computes output at each time step 

output_handwash_seir <- ode(y = state, times = times, func = handwash_I_SEIR, parms = c(beta1, beta2, gamma, omega, sigma, birth, death, p, theta))

output_handwash_seir %>% show_SEIR_model_results()

compute_cumulative_infection_time(output_handwash_seir)
```

# Basic SEIR with handwashing - affects infectious and susceptible

Model assuming handwashing impacts infectious and exposed people

```{r}
handwash_SI_SEIR<-function(t, state, parameters) {
    with(as.list(c(state, parameters)),{
        N <- S + E + I + R
        
        dW <- -theta*W + p*(N - W)
        
        dS <- -beta1*S*(1 - W/N)*(I/N) - beta2*S*(W/N)*(I/N) + birth*N - death*S + omega*R + alpha*(W/N)*E
        dE <- beta1*S*(1 - W/N)*(I/N) + beta2*S*(W/N)*(I/N) - sigma*E - death*E - alpha*(W/N)*E
        dI <- sigma*E - death*I - gamma*I
        dR <- gamma*I - death*R - omega*R
        
        # return the rates of change as a list
        list(c(dS, dE, dI, dR, dW))
    })
}
```

Parameters
```{r}
R0 <- 2.2
latent <- 3
infectious <- 7
lambda <- .2 #this is the percent by which handwashing reduces effective contact rate
p <- .2 #prob that someone washes hands
theta <- .2 #rate that handwashing wears off
alpha <- .2 #proportion of people with washed hands return to susceptible (AKA, dont become infectious)

birth <- 0
death <- 0
omega <- 0 #waning immunity
gamma <- 1/infectious #recovery rate
sigma <- 1/latent
beta1 <- R0*gamma #effective contact rate
beta2 <- (1 - lambda)*beta1

```

Run model

```{r}
state <- c(S = 999999, #population of 1,000,000, 1 person starts of infected
           E = 0,
           I = 1, 
           R = 0,
           W = 0
)

T_end <- 500 #run model for 500 time steps
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps, and computes output at each time step 

output_handwash_si_seir <- ode(y = state, times = times, func = handwash_SI_SEIR, parms = c(beta1, beta2, gamma, omega, sigma, birth, death, p, theta, alpha))

output_handwash_si_seir %>% show_SEIR_model_results()

compute_cumulative_infection_time(output_handwash_si_seir)
```


# SEIR model with two E compartments

This model assumes that once a person is exposed, they can either (1) wash hands and return to the S compartment, or (2) continue on to experience incubation in E2 and then develop an infection in I. People in I can wash hands and reduce the transmission to people they come in contact with.

```{r}
handwash_SI2E_SEIR<-function(t, state, parameters) {
    with(as.list(c(state, parameters)),{
        N <- S + E1 + E2 + I + R
        
        dW <- -theta*W + p*(N - W)
        
        dS <- -beta1*S*(1 - W/N)*(I/N) - beta2*S*(W/N)*(I/N) + birth*N - death*S + omega*R + alpha*(W/N)*E1
        dE1 <- beta1*S*(1 - W/N)*(I/N) + beta2*S*(W/N)*(I/N) - alpha*(W/N)*E1 - (1 - alpha*(W/N))*E1
        dE2 <- (1 - alpha*(W/N))*E1 - sigma*E2 - death*E2
        dI <- sigma*E2 - death*I - gamma*I
        dR <- gamma*I - death*R - omega*R
        
        # return the rates of change as a list
        list(c(dS, dE1, dE2, dI, dR, dW))
    })
}
```


Parameters
```{r}
R0 <- 2.2
latent <- 3
infectious <- 7
lambda <- .2 #this is the percent by which handwashing reduces effective contact rate
p <- .2 #prob that someone washes hands effectively
theta <- .2 #rate that handwashing wears off
alpha <- .2 #proportion of people with washed hands in E1 return to susceptible (AKA, dont become infectious)

birth <- 0
death <- 0
omega <- 0 #waning immunity
gamma <- 1/infectious #recovery rate
sigma <- 1/latent
beta1 <- R0*gamma #effective contact rate
beta2 <- (1 - lambda)*beta1
```

Run model

```{r}
show_SE2IR_model_results<-function(df1) {
    df1 %>% as_tibble() %>% mutate(S = as.numeric(S), E1 = as.numeric(E1), E2 = as.numeric(E2), I = as.numeric(I), R = as.numeric(R)) %>%
    pivot_longer(cols = c("S", "E1", "E2", "I", "R"), names_to = "compartment", values_to = "count") %>%
    mutate(compartment = factor(compartment, levels = c("S", "E1", "E2", "I", "R"))) %>% 
    ggplot(aes(x = time, y = count, color = compartment)) +
        geom_line() +
        xlab('Time') +
        ylab('Count') 
}
```


```{r}
state <- c(S = 999999, #population of 1,000,000, 1 person starts of infected
           E1 = 0,
           E2 = 0,
           I = 1, 
           R = 0,
           W = 0
)

T_end <- 500 #run model for 500 time steps
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps, and computes output at each time step 

output_handwash_si2e_seir <- ode(y = state, times = times, func = handwash_SI2E_SEIR, parms = c(beta1, beta2, gamma, omega, sigma, birth, death, p, theta, alpha))

output_handwash_si2e_seir %>% show_SE2IR_model_results()

compute_cumulative_infection_time(output_handwash_si2e_seir)
```

# Analysis

```{r}

```


Potential additions:

* Age stratification  
* Isolation upon symptom onset  
* Assumptions about number of visits out of the house per week  
* Population density (impacts beta)  

ALSO, we can do analysis based on not just CIT but also peak number of infecteds, peak number of people in the hospital, time until peak, etc.